Rhizaria in the oligotrophic ocean exhibit clear temporal and vertical variability.

### 2024-01-29 manuscript/main.qmd
render with line numbers

---
title: Rhizaria in the oligotrophic exhibit clear temporal and vertical variability.
bibliography: references.bib
csl: ./deep-sea-research-part-i.csl
execute:
  echo: false
  warning: false
  message: false
author:
  - name: Alex Barth*
    affiliations:
    - name: University of South Carolina
      department: Biological Sciences
      address: 700 Sumter St 401, Columbia, SC 29208
      email: AB93@email.sc.edu
      phone: +1 (440) 668-8376
  - name: Leo Blanco-Berical
    affiliations:
    - name: Arizone State University
  - name: Rod Johnson
    affiliations: 
    - name: Bermuda Institute of Ocean Sciences
      address: Bermuda GE 01, UK
  - name: Joshua Stone
    affiliations:
    - name: University of South Carolina
      department: Biological Sciences
      address: 700 Sumter St 401, Columbia, SC 29208
format:
  pdf:
    template-partials:
      - "before-body.tex"
      - "_titlepage.tex"
    include-in-header:
      - "in-header.tex"
    number-sections: false
    linestretch: 2
    tbl-cap-location: top
    filters: [wordcount.lua]
    keep-tex: true
---

@@

# References:
:::{#refs}
:::

>>>>

---
title: Rhizaria in the oligotrophic ocean exhibit clear temporal and vertical variability.
bibliography: references.bib
csl: ./deep-sea-research-part-i.csl
execute:
  echo: false
  warning: false
  message: false
author:
  - name: Alex Barth*
    affiliations:
    - name: University of South Carolina
      department: Biological Sciences
      address: 700 Sumter St 401, Columbia, SC 29208
      email: AB93@email.sc.edu
      phone: +1 (440) 668-8376
  - name: Leocadio Blanco-Berical
    affiliations:
    - name: Arizone State University
      department: School of Ocean Futures
      address: Bermuda Institute of Ocean Sciences 17 Biological Station, St Georges, GE 01
  - name: Rod Johnson
    affiliations: 
    - name: Arizone State University
      department: School of Ocean Futures
      address: Bermuda Institute of Ocean Sciences 17 Biological Station, St Georges, GE 01
  - name: Joshua Stone
    affiliations:
    - name: University of South Carolina
      department: Biological Sciences
      address: 700 Sumter St 401, Columbia, SC 29208
format:
  pdf:
    template-partials:
      - "before-body.tex"
      - "_titlepage.tex"
    include-in-header:
      - "in-header.tex"
    number-sections: false
    linestretch: 2
    tbl-cap-location: top
    filters: [wordcount.lua]
    keep-tex: true
abstract: |
  Recently studies have shown that Rhizaria, a super-group of marine protists have a large role in pelagic ecosystems. They are unique in that they construct mineral tests out of silica, calcium carbonate, or strontium sulfate. As a consequence, Rhizaria can have large impacts on the ocean’s cycling of carbon and other elements. However, less is known about Rhizarian ecology or their role in the pelagic food-web. Some taxa, like certain Radiolarians are mixotrophic, hosting algal symbionts. While other taxa are flux-feeders or even predatory carnivores. Some prior research has suggested that Rhizaria will partition vertically in the water column, likely due to different trophic strategies. However, very few studies have investigated their populations over extended periods of time. In this study, we present data investigating Rhizarian abundance and vertical distribution from over a year of monthly cruises into the Sargasso Sea. This study represents the first quantification of Rhizaria throughout the mesopelagic zone in an oligotrophic system for an extended period of time. We use this data to investigate the hypothesis that Rhizaria taxonomic groups will partition due to trophic mode. We also investigate how their abundance varies in accordance with a variety of environmental parameters. Rhizaria abundance was quantified using an Underwater Vision Profiler (UVP5), an in-situ imaging device. Ultimately, we show that different Rhizaria taxa will have unique vertical distribution patterns. Models relating their abundance to environmental parameters have mixed results, yet particle concentration is a common predictive variable, supporting the importance of heterotrophy amongst many taxa.
---

@@

# References:

::: {#refs}
:::

\newpage

| **Model**                          | Term                   | edf    | F     | p-value |
|-------------------|--------------|--------------|--------------|--------------|
| **All Rhizaria Epipelagic**        | Salinity               | 2.0818 | 4.654 | \<0.001 |
| $R^2_{adj}$= 0.42                  | DO                     | 0.8272 | 0.945 | 0.0136  |
|                                    | Silica                 | 3.1206 | 15.75 | \<0.001 |
|                                    | NO3                    | 2.9970 | 6.268 | \<0.001 |
|                                    | Primary Productivity   | 1.8367 | 2.099 | \<0.001 |
|                                    | Particle Concentration | 0.9282 | 2.566 | \<0.001 |
|                                    |                        |        |       |         |
| **All Rhizaria Upper Mesopelagic** | Silica                 | 0.6847 | 0.431 | 0.073   |
| $R^2_{adj}$= 0.406                 | Avg Mass Flux          | 1.5426 | 0.923 | 0.045   |
|                                    | Particle Concentration | 3.2957 | 20.89 | \<0.001 |
|                                    |                        |        |       |         |
| **All Rhizaria Lower Mesopelagic** | Avg Mass Flux          | 1.6632 | 1.824 | 0.002   |
| $R^2_{adj}$= 0.603                 | Particle Concentration | 0.7694 | 0.662 | 0.027   |

: Generalized Additive Model results for integrated total Rhizarian abundance in different regions of the water column.

\newpage

| Model                                | Term                   | edf   | F     | p-value |
|-------------------|--------------|--------------|--------------|--------------|
| **Acantharea Epipelagic**            | Salinity               | 2.264 | 4.113 | \<0.001 |
| R2adj=0.53                           | O2                     | 2.579 | 6.712 | \<0.001 |
|                                      | Avg Mass Flux          | 4.810 | 22.81 | \<0.001 |
|                                      | Bacteria #/L           | 1.599 | 1.317 | 0.0117  |
|                                      | Particle Concentration | 0.853 | 1.158 | 0.0087  |
|                                      |                        |       |       |         |
| **Acantharea Upper Mesopelagic**     | Avg C Flux             | 1.296 | 0.621 | 0.0439  |
| R2adj=0.231                          | Avg N Flux             | 1.619 | 1.283 | 0.0026  |
|                                      | Particle Concentration | 0.952 | 3.886 | \<0.001 |
|                                      |                        |       |       |         |
| **Acantharea Lower Mesopelagic**     | Temperature            | 2.076 | 4.155 | \<0.001 |
| R2adj=0.509                          | Avg N Flux             | 1.494 | 1.216 | 0.0113  |
|                                      | Primary Productivity   | 0.766 | 0.648 | 0.0253  |
|                                      | Particle Concentration | 2.037 | 10.03 | \<0.001 |
| **Aulacanthidae Epipelagic**         | Salinity               | 0.792 | 0.757 | 0.0241  |
| R2adj=0.251                          | Avg N Flux             | 0.869 | 1.324 | 0.0018  |
|                                      | Primary Productivity   | 2.312 | 7.704 | \<0.001 |
|                                      | Particle Concentration | 2.008 | 2.200 | 0.0017  |
|                                      |                        |       |       |         |
| **Aulacanthidae Upper Mesopelagic**  | Particle Concentration | 2.832 | 9.802 | \<0.001 |
| R2adj = 0.158                        |                        |       |       |         |
|                                      |                        |       |       |         |
| **Aulacanthidae Lower Mesopelagic**  | Avg Mass Flux          | 1.622 | 1.991 | 0.002   |
| R2adj=0.298                          | Particle Concentration | 2.123 | 6.706 | \<0.001 |
|                                      |                        |       |       |         |
| **Aulosphaeridae Upper Mesopelagic** | Particle Concentration | 2.653 | 13.06 | \<0.001 |
| R2adj=0.2                            |                        |       |       |         |
|                                      |                        |       |       |         |
| **Aulosphaeridae Lower Mesopelagic** | Particle Concentration | 0.972 | 6.248 | \<0.001 |
| R2adj=.147                           |                        |       |       |         |
|                                      |                        |       |       |         |
| **Castanellidae Epipelagic**         | Avg C Flux             | 1.462 | 0.816 | 0.0421  |
| R2adj = 0.124                        | Primary Productivity   | 0.771 | 0.665 | 0.0247  |
|                                      | Particle Concentration | 3.956 | 4.623 | \<0.001 |
|                                      |                        |       |       |         |
| **Coelodendridae Upper Mesopelagic** | Avg N Flux             | 0.822 | 0.919 | 0.0183  |
| R2adj=.113                           | Particle Concentration | 0.970 | 6.208 | \<0.001 |
|                                      |                        |       |       |         |
| **Coelodendridae Lower Mesopelagic** | Particle Concentration | 1.773 | 4.873 | \<0.001 |
| R2adj=.133                           |                        |       |       |         |
|                                      |                        |       |       |         |
| **Collodaria Epipelagic**            | Salinity               | 0.925 | 2.267 | \<0.001 |
| R2adj=0.16                           | O2                     | 2.015 | 2.217 | 0.002   |
|                                      | Bacteria #/L           | 1.843 | 2.100 | 0.002   |
|                                      |                        |       |       |         |
| **Foraminifera Epipelagic**          | Temperature            | 4.414 | 8.789 | \<0.001 |
| R2adj=0.445                          | O2                     | 1.603 | 1.287 | 0.0111  |
|                                      | Avg N Flux             | 2.512 | 3.396 | \<0.001 |
|                                      | Primary Productivity   | 0.824 | 0.920 | 0.0153  |
|                                      | Particle Concentration | 0.919 | 2.257 | \<0.001 |
: Taxa-specific generalized additive models for different regions of the water column.

\newpage

![Example images of different Rhizaria taxa. 4mm scale bar shown in lower right. All vignettes are same scale](images/01_taxa.pdf)

\newpage

![Environmental profiles across time-series of study period. Y axis shows depth in meters. (A) Temperature. (B) Salinity. (C) Dissolved Oxygen. (D) In-situ chlorophyll fluorescence. (E) Particle concenctrion (184 - 450$\mu m$). (F) Bacteria Abundance. (G) Silica. (H) Nitrate](images/02-environmental.pdf)

\newpage

![Primary productivity (A) and flux estimates from total mass (B), carbon (C), and nitrogen (D). Values are from monthly cruises with month displayed in corresponding colors. Absent data are shown by grey bar.](images/03_prod-flux.pdf)

\newpage

![Average abundance of Rhizaria in 25m bins, across entire study period. Shown are total Rhizaria (A), Acantharea (B), Collodaria (C), Aulacanthidae (D), Foraminifera (E), Aulosphaeridae (F), Castanellidae (G), Coelodendridae (H).](images/04_vertical.pdf)

\newpage

![Seasonality of Rhizarian integrated abundance for the epipelagic (0-200m) (A-B), upper mesopelagic (200-500m) (C-D), lower mesopelagic (500-1000m) (E-F). Left panels (A,C,E) display total integrated abundance per monthly cruise colored by month. Right panels (B, D, F) display community composition of each total abundance.](images/05_seasonality.pdf)

\newpage

![Comparison of average sizes (ESD) amongst Rhizarian taxa which occurred throughout the water column.](images/09_size-comp.pdf)

\newpage

![Partial effects of smooth terms in taxa-specific GAM models from the epipelagic (0-200m). Effects are grouped by taxa; Acantharea (A), Foraminifera (B), Aulacanthidae (C), Collodaria (D), Castanellidae (E).](images/06_epi-partials.pdf)

\newpage

![Partial effects of smooth terms in taxa-specific GAM models from the upper mesopelagic (200-500m). Effects are grouped by taxa; Acantharea (A), Aulacanthidae (B), Aulosphaeridae (C), Coelodendridae (D).](images/07_upmeso-partial.pdf)

\newpage

![Partial effects of smooth terms in taxa-specific GAM models from the lower mesopelagic (500-1000m). Effects are grouped by taxa; Acantharea (A), Aulacanthidae (B), Aulosphaeridae (C), Coelodendridae (D).](images/08_lomeso-partials.pdf)

\newpage

end.
